Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-03-30 tagless/tagless-final.tex
a few small changes

\begin{abstract}
We have built a (family of) tagless partial evaluator(s) for a typed object
language (embedded DSL), placing minimal requirements on
the type system of the meta-language:
we need let-bound polymorphism and staging, but not dependent
types, generalized algebraic data types, nor a tag elimination 
postprocessing step.

The main tool is to take functions seriously: we encode the
higher-order abstract syntax of object terms not by using constructors for
an algebraic data type but rather by invoking cogen functions, which occur as
free variables in the encoded term.  In other words, we eschew initial
algebras for the wilder world of final algebras.
Our family of evaluators contains an interpreter, a compiler, a partial
evaluator, a term-size evaluator and a CPS call-by-name transformer.
Our approach highlights the importance of order-2 type constructors
(rank-2 polymorphism), which lets us encode DSL terms with
translucent types: enough abstraction is available to interpret the
terms in various ways, and yet enought type information is available 
to typecheck the terms and statically assure their interpretation will
never get stuck.
\end{abstract}

@@

A popular way to define and implement a programming language is to embed it in
another \citep{reynolds-definitional}.  The embedded language is called the
\emph{object} or \emph{subject} language, whereas the embedding language is the
\emph{metalanguage}.  Embedding is especially appropriate for domain\hyp
specific object languages because it supports rapid prototyping and integration
with the host environment.  If the metalanguage supports \emph{staging}, then
the embedding can compile object programs to the metalanguage and avoid the
overhead of interpreting them on the fly \citep{WalidICFP02}.  A staged
definitional interpreter is thus a promising way to build a domain\hyp specific
language.

@@

We have programmed our interpreters in OCaml/MetaOCaml (\cite{metaocaml})
and Haskell. The complete code is available 
\footnote{the URL will appear in the final version of the paper.}
% connot give URL in the ICFP paper due to double-blind reviews
in the supplemental material to the paper. We will use both
(Meta)OCaml and Haskell for our examples throughout the paper.
Each language has its strength and weaknesses; this dual implementation
forced us to clarify our own ideas.  We will continually leverage the
respective strengths of these languages, as certain of our claims are
patently obvious in one implementation, while requiring more argumentation
in the other.

@@

    At POPL 2003, Taha et al~\cite{TahaPOPL03} and Xi et al~\cite{XiPOPL03}
both
interpreter of a typed language in a typed language is an interesting
problem. The known solutions include the use of the Universal type, or
GADTs. The former is not quite satisfactory for performance reasons;
and also because of the presence of partial pattern matching in the
interpreter; the inexhaustiveness of the pattern matching should never
be triggered when executing a well-typed term. Alas, this fact cannot
be made obvious to the compiler. That also exlains unoptimality.
\oleg{Should say somewhere here that our code is surprisingly simple,
  and, in the hindsight, obvious. The problem of typed direct
  intrepreter of a typed language is difficult and thought to require
  GADT or dependent types. 
  In fact, several papers introducing GADTs (Xi03 and SPJ) as well as 
  (language in Walid03) and Concoqtion, written 2003-2007, used the 
  typed tagless intrepreter as the motivating example.
  Taha (in Jones optimality paper, I think) said that typed
  interpreter of the typed language was not solved then.
  Or, should we refer to Sec `Other solutions'}

>>>>

\begin{abstract}
We have built a family of tagless embeddings (interpretations) for a
higher-order typed object
language in a typed metalanguage (current Haskell and ML), 
requiring neither dependent
types, nor generalized algebraic data types, nor a tag elimination 
postprocessing step. That has prevously considered an open problem
(impossible). 
Our family of evaluators (interpreters, embeddings?) contains 
an interpreter, a compiler, a partial
evaluator, a term-size evaluator and a CPS call-by-name transformer.
We take advantage of staging when available.

The main tool is to take functions seriously: we encode the
deBruijn or higher-order abstract syntax  of object terms 
not by using constructors for
an algebraic data type but rather by invoking cogen functions, which occur as
free variables in the encoded term.  In other words, we eschew initial
algebras for the wilder world of final algebras.
(Or: we expose the co-algebraic structure of the object language).
Our approach highlights the importance of order-2 type constructors
(rank-2 polymorphism), available with ML functors and
Haskell98 constructor classes. This lets us encode object terms with
translucent types: enough abstraction is available to interpret the
terms in various ways, and yet enought type information is available 
to typecheck the terms and statically assure their interpretation will
never get stuck.

\oleg{If we talk about self-interpreter, mention let-bound polymorphism}
\end{abstract}

@@

A popular way to define and implement a programming language is to embed it in
another \citep{reynolds-definitional}.  The embedded language is called the
\emph{object} or \emph{subject} language, whereas the host language is the
\emph{metalanguage}.  Embedding is especially appropriate for domain\hyp
specific object languages because it supports rapid prototyping and integration
with the host environment \oleg{Cite Hudaks' survey and perhaps say a
  couple of more words}.  If the metalanguage supports \emph{staging}, then
the embedding can compile object programs to the metalanguage and avoid the
overhead of interpreting them on the fly \citep{WalidICFP02}.  A staged
definitional interpreter is thus a promising way to build a domain\hyp specific
language.

@@

    At POPL 2003, Taha et al~\cite{TahaPOPL03} and Xi et al~\cite{XiPOPL03}
both
interpreter of a typed language in a typed language is an interesting
problem. The known solutions include the use of the Universal type, or
GADTs. The former is not quite satisfactory for performance reasons;
and also because of the presence of partial pattern matching in the
interpreter; the inexhaustiveness of the pattern matching should never
be triggered when executing a well-typed term. Alas, this fact cannot
be made obvious to the compiler. That also exlains unoptimality.

 The problem of typed direct
  intrepreter of a typed language is difficult and thought to require
  GADT or dependent types. 
  In fact, several papers introducing GADTs (Xi03 and SPJ) as well as 
  (language in Walid03) and Concoqtion, written 2003-2007, used the 
  typed tagless intrepreter as the motivating example.
  Taha (in Jones optimality paper, I think) said that typed
  interpreter of the typed language was not solved then.
  Or, should we refer to Sec `Other solutions'.

@@

We have programmed our interpreters in OCaml (and, for staging,
MetaOCaml (\cite{metaocaml})) and Haskell. The complete code is available 
\footnote{the URL will appear in the final version of the paper.}
% connot give URL in the ICFP paper due to double-blind reviews
in the supplemental material to the paper. We will use both
(Meta)OCaml and Haskell for our examples throughout the paper.
Each language has its strength and weaknesses; this dual implementation
forced us to clarify our own ideas.  We will continually leverage the
respective strengths of these languages, as certain of our claims are
patently obvious in one implementation, while requiring more argumentation
in the other.
Our code is surprisingly simple, and, in the hindsight, obvious. We
must emphasize however that direct typeful tagless interpretation of a typed
language until now has thought to be impossible in plain ML and
require extensions like GADT -- which we discuss in detail in Section
XXX.
